Spherical harmonic analysis of faceted spheroids identifies shaping strategies and standardisation at Qianshangying (North China)

### 2025-12-27 README.qmd
update

This project provides Python scripts for 3D spheroid shape analysis using SPHARM.  
1. **Download STL files**  
   - Go to the [OSF repository](https://osf.io/ctne9/files/osfstorage) and download the folder containing STL models: `3d_models_QSY_spheroids_multi_poly`  
2. **Unzip and place STL files in project directory**  
   - Extract the downloaded archive.  
   - After unzipping, copy the entire folder **`3d_models_QSY_spheroids_multi_poly`** into the project working directory:  [:file_folder: `SPHARM_main`](/py_SPHARM/SPHARM_main)  
3. **Open the main script and install dependencies**    
   - In [:file_folder: `SPHARM_main`](/py_SPHARM/SPHARM_main), open the main Python script file `main.py` in PyCharm.      
   - Install required dependencies by running in the Terminal:  
     ```
     python -m pip install -r requirements.txt  
     ```  
4. **Run the main script-main.py**    
   - Run `main.py` in PyCharm.    
   - The script will perform SPHARM calculation on a single 3D model and reconstruct it based on the computed results.    
   - You can save figures during the process.    
   - By default, all output results will be saved to the project folder [:file_folder: `SPHARM_main`](/py_SPHARM/SPHARM_main).    
5. **Run the batch operation script (`batch_operation.py`)**    
   - Run `batch_operation.py` in PyCharm.    
   - This script will process all STL files in the folder `3d_models_QSY_spheroids_multi_poly` and perform the following computations:    
     - SPHARM (spherical harmonics) calculation for each model    
     - Rotation-invariant spectrum computation    
     - Spherical harmonic energy (SHE) calculation    
     - UMAP analysis on the rotation-invariant spectrum    
   - You can replace the input folder with your own STL files to perform a similar analysis.    
   - By default, all output results will be saved to: [:file_folder: raw_data](/analysis/data/raw_data)    
6. **Run the curvature script (`curvature.py`)**    
   - Locate `curvature.py` in the [:file_folder: `SPHARM_module`](/py_SPHARM/SPHARM_module) and open it in PyCharm.    
   - This script will compute the curvature for all STL files in the folder `3d_models_QSY_spheroids_multi_poly`.    
   - By default, all output results will be saved to: [:file_folder: raw_data](/analysis/data/raw_data)  
7. **Run the sphericity script (`sphericity.py`)**      
   - Locate `sphericity.py` in the [:file_folder: `SPHARM_module`](/py_SPHARM/SPHARM_module) and open it in PyCharm.      
   - This script will compute the sphericity for all STL files in the folder `3d_models_QSY_spheroids_multi_poly`.      
   - By default, all output results will be saved to: [:file_folder: raw_data](/analysis/data/raw_data)

>>>>

This project provides Python scripts for 3D spheroid shape analysis using SPHARM.  
1. **Download STL files**  
   - Go to the [OSF repository](https://osf.io/ctne9/files/osfstorage) and download the folder containing STL models: `3d_models_QSY_spheroids_multi_poly`  
2. **Unzip and place STL files in project directory**  
   - Extract the downloaded archive.  
   - After unzipping, copy the entire folder **`3d_models_QSY_spheroids_multi_poly`** into the project working directory:  [:file_folder: `SPHARM_main`](/py_SPHARM/SPHARM_main)  
3. **Open the main script and install dependencies**    
   - In [:file_folder: `SPHARM_main`](/py_SPHARM/SPHARM_main), open the main Python script file `main.py` . 
   - This project was developed and tested with Python 3.10.2. To ensure reproducibility, we recommend using a Python version manager (pyenv or conda) to install the correct version. e.g.:
     ```
     pyenv install 3.10.2
     pyenv local 3.10.2  
     ``` 
   - Install required dependencies by running in the Terminal:  
     ```
     python -m pip install -r requirements.txt  
     ```  
4. **Run the main script-main.py**    
   - Run `main.py` in PyCharm.    
   - The script will perform SPHARM calculation on a single 3D model and reconstruct it based on the computed results.    
   - You can save figures during the process.    
   - By default, all output results will be saved to the project folder [:file_folder: `SPHARM_main`](/py_SPHARM/SPHARM_main).    
5. **Run the batch operation script (`batch_operation.py`)**    
   - Run `batch_operation.py` in PyCharm.    
   - This script will process all STL files in the folder `3d_models_QSY_spheroids_multi_poly` and perform the following computations:    
     - SPHARM (spherical harmonics) calculation for each model    
     - Rotation-invariant spectrum computation    
     - Spherical harmonic energy (SHE) calculation    
     - UMAP analysis on the rotation-invariant spectrum    
   - You can replace the input folder with your own STL files to perform a similar analysis.    
   - By default, all output results will be saved to: [:file_folder: raw_data](/analysis/data/raw_data)    
6. **Run the curvature script (`curvature.py`)**    
   - Locate `curvature.py` in the [:file_folder: `SPHARM_module`](/py_SPHARM/SPHARM_module) and open it in PyCharm.    
   - This script will compute the curvature for all STL files in the folder `3d_models_QSY_spheroids_multi_poly`.    
   - By default, all output results will be saved to: [:file_folder: raw_data](/analysis/data/raw_data)  
7. **Run the sphericity script (`sphericity.py`)**      
   - Locate `sphericity.py` in the [:file_folder: `SPHARM_module`](/py_SPHARM/SPHARM_module) and open it in PyCharm.      
   - This script will compute the sphericity for all STL files in the folder `3d_models_QSY_spheroids_multi_poly`.      
   - By default, all output results will be saved to: [:file_folder: raw_data](/analysis/data/raw_data)

### 2025-12-27 analysis/paper/paper.qmd
update

To address these challenges, we developed a set of quantitative methods that integrate morphological and technological analyses to investigate spheroid production strategies. While previous morphological measures have focused on the shape of spheroids close to perfect sphere, contributing to discussions of spheroid standardization, they provided limited assistance in reflecting on production strategies. We therefore adopted spherical harmonics, a landmark-free method recently introduced to archaeology. This offers morphometric precision comparable to geometric morphometrics but does not rely on predefined landmarks, making it particularly suitable for artifacts like spheroids that lack homologous landmarks. We introduce a novel modification of the spherical harmonics method to distinguish the morphological variability of flaked pieces, enabling us to differentiate between spheroids and multifacial cores or polyhedrons. To our knowledge, this is the first attempt of its kind in the field of archaeology. Additional technological analyses, including flake scar orientation analysis and reduction index calculation, further contribute to reconstructing the spheroid production process. All of the methods we present here are open-source to enable replication of our results and support unified, cross-regional quantitative studies in the future.

@@

A total of 956 stone artifacts were recovered from four localities at Qianshangying. Following a widely used techno-typological framework [@Isaac1981; @Isaac1986; @delatorre2004; @de2018oldowan; @pei2017] , the assemblage was classified into four general categories: flaked pieces, detached pieces, pounded pieces, and unmodified materials. The numerical distribution and proportional composition of these categories are provided in the supplementary material. Within the category of flaked pieces, 6 spheroids and 7 subspheroids were identified; together forming a spheroid group of 13 artifacts, accounting for 1.3% of the total lithic assemblage. They are all faceted stone tools with intensive flake scars on the surface. Within this continuum, spheroids show a more symmetrical morphology and more extensive flake scar coverage, while subspheroids are characterized by lower symmetry and a less intensive degree of shaping. Also within the category of flaked pieces, multifacial cores and polyhedrons are intensively flaked cores with more than three flaking surfaces, with polyhedrons generally regarded as approaching spherical forms. However, using spherical forms as the distinguish standard is still subjective. In this study, artifacts with a thickness-to-length ratio greater than 0.6 are operationally classified as polyhedrons (n = 17), as these specimens tend to exhibit more comparable dimensions along the three principal axes, whereas the remaining pieces are classified as multifacial cores (n = 12). Importantly, this classification reflects a morphological continuum rather than discrete categories; the threshold value of 0.6 is adopted as a practical criterion to distinguish polyhedrons from multifacial cores, producing roughly comparable sample sizes for the two categories within the assemblage.

@@

This method has been widely used in the field of biology, for example in the analysis of skull and cell morphology [@Link2024; @Medyukhina2020; @Harper2022; @Hewitt2024; @Grieb2022]. However, there have been very few applications of SPHARM to archaeological artifacts. @sholts2017tracing applied SPHARM-PDM to standardize the number and position of vertices across early North American bifaces (ca 13,100–9,000 BP), enabling point-by-point comparison of original and mirrored shapes to assess asymmetry variation, but SPHARM was not directly used to describe stone tool shape. @Muller2023 used closed-source and proprietary Matlab software to investigate intentionality in the production of 150 spheroids from 'Ubeidiya (ca 1.4 Ma) in the Levant. @Muller2023 did not provide code to document their analysis, but they found that lithic reduction resulted in the spheroids becoming more spherical, and concluded that spheroids represent intentionally knapped items. @Noshita2025 used 3D Slicer-SlicerSALT to apply spherical harmonic analysis to Ongagawa-style pottery in the Yayoi period of Japan (800 BC–AD 250). Their results were consistent with the hypothesis of an association of Ongagawa-style pottery with the spread of agriculture via two routes in prehistoric Japan, and they provided code and date files to enable their analysis to be reproduced.

Our work combines the merits of these previous studies to demonstrate an open-source and reproducible analysis of stone artefacts using SPHARM. We further introduce the use of rotation-invariant power spectrum and UMAP to extract shape features, explore hidden structures in data, and then achieve grouping among spheroids and cores, marking a first in archaeological research while substantially extending the depth of analysis provided by SPHARM. All these steps were completed in Python with the code openly available online at https://osf.io/ctne9/. Our work demonstrates the value of this method for quantitative shape analyses of objects that are roughly spherical in shape, such as the spheroids and cores in our case study from Qianshangying.

@@

As shown in the @fig-SPHARM-steps, following the approach by @Medyukhina2020 and @Link2024, we first imported, cleaned, simplified and normalized the STL files of spheroids and cores to obtain 3D mesh vertices, which were then converted from Cartesian (x, y, z) to spherical coordinates (r, θ, φ). Next, we interpolated radial values onto a regular spherical grid of N×N and then decomposed them using spherical harmonics based on the Driscoll–Healy sampling scheme [@wieczorek2018shtools]. The resulting coefficients were normalized for scale-invariant comparison across models.

From the normalized spherical harmonic coefficients, we computed the rotation-invariant power spectrum by summing the squared magnitudes of the coefficients at each degree. Each 3D shape is thus represented by its unique power spectrum, which serves as a multidimensional descriptor. To explore relationships between models, we applied a Uniform Manifold Approximation and Projection (UMAP) for dimensionality reduction and visualization [@Van2023]. Unlike traditional linear dimensionality reduction methods such as Principal Component Analysis (PCA), which capture only global linear variance patterns and typically rely on the first few spherical harmonic coefficients, UMAP is a non-linear technique designed to preserve both local and global data structure. Its grouping is based on the local and global similarity of shape data. This makes UMAP particularly well suited for complex shape data represented by spherical harmonic power spectrum, where morphological variations may manifest in subtle, non-linear ways across multiple degrees. By effectively capturing these non-linear relationships, UMAP yields a refined morphospace representation, enabling detection of complex shape differences that linear techniques may overlook.

To assist in the interpretation of the groupings revealed by UMAP, we included a set of standard geometric models—sphere, rounded cube, ellipsoid (1.2:1:0.8), rounded box (1.2:1:1), and disc-like forms (flattened and discoid core)—in the SPHARM analysis and UMAP embedding as visual references. Spheroids, polyhedrons and multifacial cores were positioned in this morphospace based on their morphological similarity to these reference models: individuals that cluster more closely to a specific standard form exhibit corresponding shape characteristics.

@@

# Mann-Whitney U test
# between sphericity_poly_multi and sphericity_sub_spheroid

group_poly_multi <- mean_sphericity_tbl$Sphericity[mean_sphericity_tbl$Group == "Polyhedron + Multifacial"]
group_sub_spheroid <- mean_sphericity_tbl$Sphericity[mean_sphericity_tbl$Group == "Spheroid + Subspheroid"]

@@

@@

umap_data <- read_csv(here("analysis/data/derived_data/umap_lmax20.csv"))

@@

# Elbow method for optimal K
#wss <- numeric(10)

@@

ggsave(here("analysis/paper/figures/p_umap_by_category.png"),
       plot = p_umap_by_cluster_with_images,
       width = 10, height = 6, dpi = 300)

@@

@tbl-1 summarizes the basic information (rock type, blank type, final morphology, and cortex coverage) and morphological analysis results (classifications derived from SPHARM, values for sphericity and spherical harmonic energy) of the spheroids, subspheroids, polyhedrons and multifical cores involved in the research. From this table, we can observe distinctions between spheroids and subspheroids (n = `r n_sub_spheroid`) on the one hand, and polyhedrons and multifacial cores (n =`r n_poly_multi`) on the other. In the subsequent analysis, these categories are treated as two separate groups for comparison.

Our sphericity results reveals a clear difference between the group of spheroids and the group of polyhedrons and multifacial cores. The mean sphericity of spheroids (`r mean_sphericity_sub_spheroid`) is notably higher than that of polyhedrons and multifacial cores (`r mean_sphericity_poly_multi`), with a Wilcoxon rank-sum test yielding W = `r U_value`, p = $`r p_value_Mann_sci`$. This result indicates that spheroids possess a more spherical geometry overall, supporting their typological classification as a distinct type.

@@

file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape.xlsx")
data_raw2 <- 
  read_excel(file_path) %>% 
  drop_na() %>% 
  mutate(across(where(is.numeric), \(x) round(x, 3)))

@@

```{r, echo=FALSE}
#| label: fig-spearman-correlation-matrix
#| fig-cap: "Spearman correlation matrix of shape variables. Red star indicates p < 0.05"

file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape.xlsx")
data <- read_excel(file_path)

colnames(data) <- make.names(colnames(data))

selected_names <- c("Scar.number", "Removal.ratio", "SDI",
                    "Edge.angle.mean", "Curvature", "Sphericity", "SHE")

selected_cols <- data[, selected_names[selected_names %in% colnames(data)]]

# print(dim(selected_cols))
# print(colnames(selected_cols))

file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape.xlsx")
data <- read_excel(file_path)
colnames(data) <- make.names(colnames(data))

@@

p <- 
  ggplot(labels_long, aes(x = Var2_label, y = Var1_label, fill = cor)) +
    geom_tile(color = "gray") +
    scale_fill_gradient2(low = "#5e3c99", mid = "white", high = "#e66101", midpoint = 0) +
    geom_text(aes(label = cor_label), size = 3) +
    geom_text(aes(label = star), color = "red", size = 6, nudge_y = 0.25) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(fill = "Correlation", x = "", y = "")
p

#ggsave(here("analysis/paper/figures/spearman_correlation_matrix.png"),
       #plot = p, width = 8, height = 6, dpi = 300)
```

For all spheroids and subspheroids, we conducted shaping efficiency analysis. @fig-spearman-correlation-matrix shows significant positive correlations among the three variables representing reduction intensity: scar number, removal ratio, and SDI. Among them, scar counts show significant positive correlations with both removal ratio (ρ = `r cor_scar_removal`, p = $`r p_scar_removal`$) and SDI (ρ = `r cor_scar_sdi`, p = $`r p_scar_sdi`$). Removal ratio and SDI exhibit a stronger and more highly significant correlation (ρ = `r cor_removal_sdi`, p = $`r p_removal_sdi`$). These two variables incorporate surface area in their calculation, making them more effective in capturing the density and intensity of reduction, and therefore more reliable indicators of the shaping process in spheroids.

Among the shape descriptors, we observed several significant correlations. Mean edge angle was positively correlated with sphericity (ρ = `r cor_angle_sph`, p = $`r p_angle_sph`$) and negatively correlated with Spherical Harmonics Power (SHE) (ρ = `r cor_angle_she`, p = $`r p_angle_she`$), indicating that greater edge angles are associated with more spherical and regular forms. As noted in previous studies, mean edge angle is a key variable in capturing the morphology of spheroids [@tittonSubspheroidsLithicAssemblage2020]. This correlation between edge angle and SHE may help explain why spheroids were often interpreted as exhausted cores in earlier research [@sahnouni1997]. Indeed, spheroids must be significantly reduced, as achieving a near-perfect spherical shape typically results in increased edge angles, making further flake removal difficult. However, the reverse is not necessarily true, as not all exhausted cores are shaped like spheroids.

Curvature was only significantly correlated with sphericity (ρ = `r cor_cur_sph`, p = $`r p_cur_sph`$), suggesting that smoother surfaces tend to approximate spheres more closely. Its lack of correlation with other variables may reflect its focus on local surface texture, while edge angle and SHE capture broader aspects of shape regularity and symmetry. The weak correlation between sphericity and SHE may result from the small sample size or their different sensitivities: sphericity, based on area and volume, tolerates surface variation, while SHE responds more to asymmetry and shape irregularity. Thus, even in specimens with similar sphericity, surface differences can produce substantial variation in SHE. The analysis of spheroid shape variables indicates that mean edge angle, curvature, sphericity, and SHE each capture distinct aspects of spheroid morphology, which explains the varying strengths of correlation observed among them.

Significant correlations were found between reduction intensity and mean edge angle and SHE. Mean edge angle was positively correlated with both scar number (ρ = `r cor_edge_scar`, p = $`r p_edge_scar`$) and removal ratio (ρ = `r cor_edge_removal`, p = $`r p_edge_removal`$), while SHE was negatively correlated with removal ratio (ρ = `r cor_she_removal`, $`r p_she_removal`$) and SDI (ρ = `r cor_she_sdi`, p = $`r p_she_sdi`$). These results suggest that as reduction progresses, spheroids develop larger edge angles and lower SHE, indicating increased regular forms and global symmetry.

@@

Overall, the observed relationships between reduction intensity and shape descriptors, particularly mean edge angle and SHE, support the interpretation that spheroid shaping was an efficient and purposeful process. The goal was to produce well-structured, globally spherical spheroids rather than smooth objects.

>>>>

To address these challenges, we developed a set of quantitative methods that integrate morphological and technological analyses to investigate spheroid production strategies. While previous morphological measures have focused on the shape of spheroids close to perfect sphere, contributing to discussions of spheroid standardization, they provided limited assistance in reflecting on production strategies. We therefore adopted spherical harmonics, a landmark-free method recently introduced to archaeology. This offers morphometric precision comparable to geometric morphometrics but does not rely on predefined landmarks, making it particularly suitable for artifacts like spheroids that lack homologous landmarks. Compared to previous studies, we employed rotation-invariant spectrum and UMAP to analyze spherical harmonics results, distinguishing the morphological variability of flaked pieces and thereby differentiating spheroids from intensively flaked cores, while extending the application of spherical harmonics. Additional technological analyses, including flake scar orientation analysis and reduction index calculation, further contribute to reconstructing the spheroid production process. All of the methods we present here are open-source to enable replication of our results and support unified, cross-regional quantitative studies in the future.

@@

A total of 956 stone artifacts were recovered from four localities at Qianshangying. Following a widely used techno-typological framework [@Isaac1981; @Isaac1986; @delatorre2004; @de2018oldowan; @pei2017] , the assemblage was classified into four general categories: flaked pieces, detached pieces, pounded pieces, and unmodified materials. The numerical distribution and proportional composition of these categories are provided in the supplementary material. The distinction between spheroids and other lithic categories was independently assessed by two researchers. Spheroids and subspheroids are all faceted stone tools with intensive flake scars on the surface. Within this continuum, spheroids show a more symmetrical morphology and more extensive flake scar coverage, while subspheroids are characterized by lower symmetry and a less intensive degree of shaping. Compared to multifacial cores and polyhedrons, spheroids and subspheroids are characterized by more rounded edges rather than sharp angles, consistent with our consensus definition of near-spherical forms. Spheroids were confirmed by both researchers and classified as spheroids; together forming a spheroid group of 13 artifacts, including 6 spheroids and 7 subspheroids, accounting for 1.3% of the total lithic assemblage. Multifacial cores and polyhedrons are intensively flaked cores with more than three flaking surfaces, with polyhedrons generally regarded as approaching spherical forms [@de2011early]. In this study, artifacts with a thickness-to-length ratio greater than 0.6 are operationally classified as polyhedrons (n = 17), as these specimens tend to exhibit more comparable dimensions along the three principal axes, whereas the remaining pieces are classified as multifacial cores (n = 12). Importantly, this classification reflects a morphological continuum rather than discrete categories; the threshold value of 0.6 is adopted as a practical criterion to distinguish polyhedrons from multifacial cores, producing roughly comparable sample sizes for the two categories within the assemblage.

@@

This method has been widely used in the field of biology, for example in the analysis of skull and cell morphology [@Link2024; @Medyukhina2020; @Harper2022; @Hewitt2024; @Grieb2022]. In contrast, archaeological applications of SPHARM remain limited and have followed a small number of distinct methodological pathways, including the use of SPHARM-based point distribution models to establish surface correspondence for asymmetry assessment (e.g. @sholts2017tracing), the application of spherical harmonics to test hypotheses of intentionality in lithic reduction (e.g. @Muller2023), and the construction of empirical morphospaces via PCA of SPHARM coefficients for descriptive shape analysis (e.g. @Noshita2025). Despite demonstrating the potential of SPHARM for archaeological materials, these studies commonly focus on descriptive interpretations of spherical harmonic coefficients or their principal components, and are largely confined to linear dimensionality reduction frameworks, which may limit the exploration of high-resolution or non-linear structure in shape variation. Moreover, open and fully reproducible analytical workflows are still rarely implemented in archaeological applications of SPHARM.

Our work builds on previous archaeological applications of SPHARM to present an open-source and fully reproducible workflow for quantitative shape analysis of stone artefacts. We extend existing approaches by combining rotation-invariant spherical harmonic power spectrum with non-linear dimensionality reduction using Uniform Manifold Approximation and Projection (UMAP), enabling feature-based exploration and unsupervised grouping of spheroids and cores, which substantially extends the depth of analysis provided by SPHARM. All these steps were completed in Python 3.10.2 with the code openly available online at https://osf.io/ctne9/. Here, We use spheroids from Qianshangying as a case study to illustrate the workflow.

@@

As shown in @fig-SPHARM-steps, following @Medyukhina2020 and @Link2024, scanned models of spheroids and cores were imported, cleaned, simplified, and normalized to obtain 3D mesh vertices. These vertices were converted from Cartesian (x, y, z) to spherical coordinates (r, θ, φ), and radial values were interpolated onto a regular N×N spherical grid. Spherical harmonic decomposition was then performed using the Driscoll–Healy sampling scheme [@wieczorek2018shtools], and the resulting coefficients were normalized to allow scale-invariant comparison across models.

From the normalized spherical harmonic coefficients, we computed the rotation-invariant power spectrum by summing the squared magnitudes of coefficients at each degree, yielding a multidimensional and orientation-independent shape descriptor for each model. Using these rotation-invariant power spectra as input, we applied UMAP for non-linear dimensionality reduction and visualization [@Van2023], preserving both local and global similarity structure in the shape data. Clusters in the resulting UMAP space were identified using K-means clustering, providing a quantitative basis for grouping spheroids, polyhedrons, and multifacial cores. To aid interpretation of the morphospace and cluster assignments, a set of standard geometric reference models (sphere, rounded cube, ellipsoid, rounded box, and disc-like forms) were included in the analyses, allowing artefacts to be positioned relative to these reference shapes and facilitating morphological interpretation of groupings based on both cluster membership and shape similarity.

@@

# Mann-Whitney U test
# between sphericity_poly_multi and sphericity_sub_spheroid
str(mean_sphericity_tbl)
group_poly_multi <- mean_sphericity_tbl$Sphericity[mean_sphericity_tbl$Group == "Polyhedron + Multifacial"]
group_sub_spheroid <- mean_sphericity_tbl$Sphericity[mean_sphericity_tbl$Group == "Spheroid + Subspheroid"]

@@

# 从 data_clean 中直接提取三组数据
group_multifacial <- data_clean$Sphericity[data_clean$Typology == "Multifacial"]
group_polyhedron <- data_clean$Sphericity[data_clean$Typology == "Polyhedron"]
group_sub_spheroid <- data_clean$Sphericity[data_clean$Typology %in% c("Spheroid", "Subspheroid")]

# 检查数据
cat("=== 数据检查 ===\n")
cat("Multifacial 样本量:", length(group_multifacial[!is.na(group_multifacial)]), "\n")
cat("Polyhedron 样本量:", length(group_polyhedron[!is.na(group_polyhedron)]), "\n")
cat("Spheroid/Subspheroid 样本量:", length(group_sub_spheroid[!is.na(group_sub_spheroid)]), "\n\n")

# Mann-Whitney U检验1: Multifacial vs Spheroids/Subspheroids
if (length(group_multifacial[!is.na(group_multifacial)]) > 0 && 
    length(group_sub_spheroid[!is.na(group_sub_spheroid)]) > 0) {

  cat("=== Multifacial vs Spheroids/Subspheroids ===\n")
  mann_whitney_multifacial <- wilcox.test(
    group_multifacial[!is.na(group_multifacial)], 
    group_sub_spheroid[!is.na(group_sub_spheroid)], 
    exact = FALSE
  )
  print(mann_whitney_multifacial)
  cat("\n")

} else {
  cat("Multifacial vs Spheroids/Subspheroids: 数据不足\n\n")
}

# Mann-Whitney U检验2: Polyhedron vs Spheroids/Subspheroids
if (length(group_polyhedron[!is.na(group_polyhedron)]) > 0 && 
    length(group_sub_spheroid[!is.na(group_sub_spheroid)]) > 0) {

  cat("=== Polyhedron vs Spheroids/Subspheroids ===\n")
  mann_whitney_polyhedron <- wilcox.test(
    group_polyhedron[!is.na(group_polyhedron)], 
    group_sub_spheroid[!is.na(group_sub_spheroid)], 
    exact = FALSE
  )
  print(mann_whitney_polyhedron)

} else {
  cat("Polyhedron vs Spheroids/Subspheroids: 数据不足\n")
}

# Mann-Whitney U检验3: Multifacial vs Polyhedron
if (length(group_multifacial[!is.na(group_multifacial)]) > 0 && 
    length(group_polyhedron[!is.na(group_polyhedron)]) > 0) {

  cat("\n=== Multifacial vs Polyhedron ===\n")
  mann_whitney_multi_poly <- wilcox.test(
    group_multifacial[!is.na(group_multifacial)], 
    group_polyhedron[!is.na(group_polyhedron)], 
    exact = FALSE
  )
  print(mann_whitney_multi_poly)

} else {
  cat("\nMultifacial vs Polyhedron: 数据不足\n")
}

U_multi_poly <- mann_whitney_multi_poly$statistic  
p_multi_poly <- mann_whitney_multi_poly$p.value

p_multi_poly_sci <- pretty_print_sci(p_multi_poly)

@@

umap_data <- read_csv(here("analysis/data/derived_data/umap_lmax20-0.csv"))

@@

umap_data %>%
  count(category_lower)
# Elbow method for optimal K
#wss <- numeric(10)

@@

ggsave(here("analysis/paper/figures/p_umap_by_category1.png"),
       plot = p_umap_by_cluster_with_images,
       width = 10, height = 6, dpi = 300)

@@

@tbl-1 summarizes the basic information (rock type, blank type, final morphology, and cortex coverage) and morphological analysis results (classifications derived from SPHARM, values for sphericity and spherical harmonic energy) of the spheroids, subspheroids, polyhedrons and multifacial cores involved in the research.

Our sphericity results reveals no significant difference between multifacial cores and polyhedrons(a Wilcoxon rank-sum test yielding W = `r U_multi_poly`, p = $`r p_multi_poly_sci`$). Accordingly, for subsequent analysis, we combined these two into a single group, which is then compared with the group of spheroids and subspheroids. A clear difference is shown between the group of spheroids and the group of polyhedrons and multifacial cores. The mean sphericity of spheroids (`r mean_sphericity_sub_spheroid`) is notably higher than that of polyhedrons and multifacial cores (`r mean_sphericity_poly_multi`), with a Wilcoxon rank-sum test yielding W = `r U_value`, p = $`r p_value_Mann_sci`$. This result indicates that spheroids possess a more spherical geometry overall, supporting their typological classification as a distinct type.

@@

file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape_spheroid.xlsx")
data_raw2 <- 
  read_excel(file_path) %>% 
  drop_na() %>% 
  mutate(across(where(is.numeric), \(x) round(x, 3)))

@@

```{r, echo=FALSE}
#| label: fig-spearman-correlation-matrix
#| fig-cap: "Spearman correlation matrix of shape variables. Left: Spheroids and subspheroids. Right: Polyhedrons and multifacial cores. Circle indicates p < 0.05, with size representing correlation magnitude. Non-significant correlations (p ≥ 0.05) are shown without circles"

library(png)

file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape_spheroid.xlsx")
data <- read_excel(file_path)
colnames(data) <- make.names(colnames(data))

@@

p_combined <- plot_grid(
  ggdraw() + draw_image(here("analysis/paper/figures/spearman_correlation_matrix_spheroid.png")),
  ggdraw() + draw_image(here("analysis/paper/figures/spearman_correlation_matrix_core.png")),
  ncol = 2  
)

# 保存组合图像
output_dir <- here("analysis", "paper", "figures")
if(!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

ggsave(
  plot = p_combined,
  filename = here("analysis/paper/figures/combined_spearman_correlation_matrices.png"),
  width = 12,
  height = 6,
  dpi = 300,
  bg = "white"
)

# 在文档中显示图形
knitr::include_graphics(here("analysis/paper/figures/combined_spearman_correlation_matrices.png"))

```{r, echo=FALSE}
file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape_spheroid.xlsx")
data <- read_excel(file_path)
colnames(data) <- make.names(colnames(data))

# 使用corrplot包绘制相关性矩阵
library(corrplot)

# 设置颜色渐变
my_colors <- colorRampPalette(c("#5e3c99", "white", "#e66101"))(100)

# 创建corrplot图形
# 保存为PNG
png(here("analysis/paper/figures/spearman_correlation_matrix.png"), 
    width = 8, height = 6, units = "in", res = 300)
corrplot(cor_matrix, 
         method = "circle",
         type = "upper",
         order = "original",
         tl.col = "black",
         tl.srt = 45,
         diag = FALSE,
         col = my_colors,
         addCoef.col = "black",
         number.cex = 0.7,
         p.mat = p_matrix,
         insig = "blank",
         sig.level = 0.05,
         mar = c(0, 0, 0, 0),
         cl.pos = "r")
dev.off()

```{r, echo=FALSE}
file_path <- here("analysis/data/derived_data/flaking_intensity_and_shape_core.xlsx")
data <- read_excel(file_path)
colnames(data) <- make.names(colnames(data))

# 使用corrplot包绘制相关性矩阵
library(corrplot)

# 设置颜色渐变
my_colors <- colorRampPalette(c("#5e3c99", "white", "#e66101"))(100)

# 创建corrplot图形
# 保存为PNG
png(here("analysis/paper/figures/spearman_correlation_matrix2.png"), 
    width = 8, height = 6, units = "in", res = 300)
corrplot(cor_matrix, 
         method = "circle",
         type = "upper",
         order = "original",
         tl.col = "black",
         tl.srt = 45,
         diag = FALSE,
         col = my_colors,
         addCoef.col = "black",
         number.cex = 0.7,
         p.mat = p_matrix,
         insig = "blank",
         sig.level = 0.05,
         mar = c(0, 0, 0, 0),
         cl.pos = "r")
dev.off()

For group of spheroids and subspheroids, we conducted shaping efficiency analysis.The same analysis on group of polyhedrons and multifacial cores as a comparison.  @fig-spearman-correlation-matrix shows the results. Among reduction intensity descriptors, removal ratio and SDI incorporate surface area in their calculation, making them more effective in capturing the density and intensity of reduction, and therefore more reliable indicators of the reduction process compared to scar number.

Among the shape descriptors, we observed several significant correlations in the group of spheroids and subspheroids. Mean edge angle was positively correlated with sphericity (ρ = `r cor_angle_sph`, p = $`r p_angle_sph`$) and negatively correlated with Spherical Harmonics Power (SHE) (ρ = `r cor_angle_she`, p = $`r p_angle_she`$), indicating that greater edge angles are associated with more spherical and regular forms. Curvature was only significantly correlated with sphericity (ρ = `r cor_cur_sph`, p = $`r p_cur_sph`$), suggesting that smoother surfaces tend to approximate spheres more closely. Its lack of correlation with other variables may reflect its focus on local surface texture, while edge angle and SHE capture broader aspects of shape regularity and symmetry. The weak correlation between sphericity and SHE may result from the small sample size or their different sensitivities: sphericity, based on area and volume, tolerates surface variation, while SHE responds more to asymmetry and shape irregularity. Thus, even in specimens with similar sphericity, surface differences can produce substantial variation in SHE. The analysis of spheroid shape variables indicates that mean edge angle, curvature, sphericity, and SHE each capture distinct aspects of spheroid morphology, which explains the varying strengths of correlation observed among them.

Significant correlations were found between reduction intensity and mean edge angle and SHE in the group of spheroids and subspheroids. Mean edge angle was positively correlated with both scar number (ρ = `r cor_edge_scar`, p = $`r p_edge_scar`$) and removal ratio (ρ = `r cor_edge_removal`, p = $`r p_edge_removal`$), while SHE was negatively correlated with removal ratio (ρ = `r cor_she_removal`, $`r p_she_removal`$) and SDI (ρ = `r cor_she_sdi`, p = $`r p_she_sdi`$). These results suggest that as reduction progresses, spheroids develop larger edge angles and lower SHE, indicating increased regular forms and global symmetry.

@@

In contrast, the polyhedron and multifacial cores group exhibited a distinct morphological trajectory: higher core reduction intensity correlated with lower sphericity and higher SHE values. This pattern suggests that sustained flaking activity in polyhedral cores leads to increasingly irregular forms that deviate from spherical geometry. This comparative analysis allows us to address our initial research question: spheroids do not represent a product of intensive flaking in the same manner as polyhedrons. Indeed, while spheroids must undergo significant reduction to achieve their near-spherical morphology, conversely, not all exhausted cores acquire spherical morphology.

Overall, the observed relationships between reduction intensity and shape descriptors support the interpretation that spheroid shaping was an efficient and purposeful process. The goal was to produce well-structured, globally spherical spheroids rather than smooth objects.
